---
title: "eurodata -- R package for fast and easy Eurostat data import and search"
author: "Aleksander Rutkowski"
output: github_document
always_allow_html: yes
---

```{r helpers, include=FALSE}
options(width = 120)
cran <- function(package)
    paste0('https://CRAN.R-project.org/package=',package)
rdoc <- function(package,funct)
    paste0('https://www.rdocumentation.org/packages/',
           package,'/functions/',funct,'/')
lnk_ <- function(name,url)
    paste0('[',name,']','(',url,')')
lnk <- function(package,funct)
    paste0(lnk_(package,cran(package)),'::',
           lnk_(funct,rdoc(package,funct)))
estat_toc_link <-
    'https://ec.europa.eu/eurostat/api/dissemination/catalogue/toc/txt?lang=EN'
```

#### *NEW (May 2023):*

The package is fully compatible with the new [Eurostat’s API SDMX 2.1](https://wikis.ec.europa.eu/display/EUROSTATHELP/Transition+-+from+Eurostat+Bulk+Download+to+API)
and no longer relies on the old Eurostat’s Bulk Download Facility.

#### NEW (January 2023):

**_Use [eurodata_codegen](https://github.com/alekrutkowski/eurodata_codegen), a point-and-click app
for rapid and easy generation of richly-commented R code, to import a Eurostat dataset or it's subset
(based on the `eurodata::importData()` function)._**

The core API of the `eurodata` package contains just 6 functions -- 4 for data
or metadata imports and 2 for search:

Import functionality:

- **importData** -- fast thanks to `r lnk('data.table','fread')`
- **importDataLabels** -- as above
- **importMetabase** -- as above
- **importDataList** -- reflects the hierarchical structure of the Eurostat tree of datasets --
fast transformation of the raw [Table of Contents file](`r estat_toc_link`)
is based on a C++ code snippet compiled via [Rcpp](`r cran('Rcpp')`)

Search functionality:

- **browseDataList** -- based on importDataList, shows an HTML table
(generated with `r lnk('xtable','xtable')`) in a browser with a list of the found datasets
- **find** -- based on importDataList, shows a textual report on the found datasets --
a ``quick-n-dirty'' way to find a Eurostat dataset without much typing (with a keyword or a few keywords)

#### NEW (December 2022):

Parameter `filters` in `importData()` allows to download only the selected values of dimensions,
instead of downloading the full dataset. See the example close to the end below.

#### NEW (August 2022) Extra functionality:

- **describe** -- describe a given Eurostat dataset on the basis of information from Metabase
- **compare** -- compare specific Eurostat datasets on the basis of information from Metabase

See the usage example at the very end below.

## Installation

```{r installation, eval=FALSE, include=TRUE}
install.packages('eurodata') # from CRAN
# or
remotes::install_github('alekrutkowski/eurodata') # package 'remotes' needs to be installed
```

## Functionality demo

```{r attach_library}
library(eurodata)
```

### Imports

```{r import_demo, warning=FALSE, , message=FALSE}
x <- importData('nama_10_a10')  # actual dataset
str(x)
head(x,10)
y <- importDataList()  # metadata
colnames(y)
str(y[y$Code=='nama_10_a10',])  # metadata on x
z <- importLabels('geo')
head(z,10)
```

### Search

```{r background_setup, include=FALSE}
# A trick to display the side-effects:
library(magrittr)
cleanForGH <- function(x) x %>%
    substr(457, nchar(.)) %>%  # drop html style preamble not supported by GitHub md parser
    paste('<html>',.) %>%
    cat(sep='\n')
browseDataList <- function(...) eurodata::browseDataList(...)$html %>%
    cleanForGH
print.EurostatDataList <- function(x, SearchCriteria, ...) x %>%
    eurodata:::tableToHtml(Sys.time(), SearchCriteria) %>%
    list(html=.) %>%
    eurodata:::print.BrowsedEurostatDatasetList() %>%
    cleanForGH
find <- function(...) eurodata::find(...)$report %>%
    cat(sep='\n')
```

```{r find_demo}
# Free-style text search based on the parts of words in the dataset names
find(gdp,main,international,-quarterly)
# Search based on the parts of the dataset codes
find(bop, its)
find(bop,-ybk,its)
```

```{r browse_demo, results='asis'}
browseDataList(grepl('GDP',`Dataset name`) &
                   grepl('main',`Dataset name`) &
                   grepl('international',`Dataset name`) &
                   !grepl('quarterly',`Dataset name`))
browseDataList(grepl('bop',Code) & grepl('its',Code))
```

```{r}
# Producing a table of datasets which (1) include a dimension `sizeclas`
# (i.e. firm size class) and (2) some data for firms with fewer than 10 employees
# (`sizeclas` code "LT10") and (3) have sectorial data (i.e. include a
# dimension `nace_r2`).
library(magrittr)
metab <- importMetabase()
codes_with_nace <- metab %>% 
    subset(Dim_name=='nace_r2') %>%
    extract2('Code') %>%
    unique
final_codes <- metab %>%
    subset(Dim_name=='sizeclas' & Dim_val=='LT10' &
               Code %in% codes_with_nace) %>%
    extract2('Code') %>%
    unique
```

```{r, results='asis'}
importDataList() %>%
    subset(Code %in% final_codes) %>%
    as.EurostatDataList %>%
    # the `SearchCriteria` argument below is optional
    print(SearchCriteria =
              'those including data on firms with fewer than 10 employees and NACE Rev.2 disaggregation') 
```

## New parameter `filters`

To reduce the download size and time if full dataset not needed, e.g.:

```{r}
subset_of__bop_its6_det <-
    importData('bop_its6_det',
               # New -- download only subset of available data
               filters = list(geo=c('AT','BG'), # these two countries
                              TIME_PERIOD=2014:2020, # only that period
                              bop_item='SC')) # only "Services: Transport"
str(subset_of__bop_its6_det)
```

## Extras

```{r}
describe('nama_10_gdp')
describe('nama_10_gdp', wide=TRUE)
compare('nama_10_gdp', 'nama_10_a64')
```
