---
title: "eurodata -- R package for fast and easy Eurostata data import and search"
author: "Aleksander Rutkowski"
output:
  html_document:
    keep_md: true
---

<style type="text/css">
.main-container {
  max-width: 1280px;
}
</style>

```{r helpers, include=FALSE}
cran <- function(package)
    paste0('https://cran.r-project.org/web/packages/',
           package,'/index.html')
rdoc <- function(package,funct)
    paste0('http://www.rdocumentation.org/packages/',
           package,'/functions/',funct)
lnk_ <- function(name,url)
    paste0('[',name,']','(',url,')')
lnk <- function(package,funct)
    paste0(lnk_(package,cran(package)),'::',
           lnk_(funct,rdoc(package,funct)))
estat_toc_link <-
    'http://ec.europa.eu/eurostat/estat-navtree-portlet-prod/BulkDownloadListing?file=table_of_contents_en.txt'
```

The package relies on [Eurostat's Bulk Download Facility](http://ec.europa.eu/eurostat/data/bulkdownload).

The API contains just 5 functions -- 3 for data or metadata imports and 2 for search:

Import functionality:

- **importData** -- fast thanks to `r lnk('data.table','fread')`
- **importDataLabels** -- as above
- **importDataList** -- reflects the hierarchical structure of the Eurostat tree of datasets --
fast transformation of the raw [Table of Contents file](`r estat_toc_link`)
is based on a C++ code snipped compiled via [Rcpp](`r cran('Rcpp')`)  

Search functionality:

- **browseDataList** -- based on importDataList, shows an HTML table
(generated with `r lnk('xtable','xtable')`) in a browser with a list of the found datasets
- **find** -- based on importDataList, shows a textal report on the found datasets --
a ``quick-n-dirty'' way to find a Eurostat dataset witout much typing (with a keyword or a few keywords)

## Installation

```{r installation, eval=FALSE, include=TRUE}
devtools::install_github('alekrutkowski/eurodata') # package 'devtools' needs to be installed
```

## Functionality demo

```{r attach_library}
library(eurodata)
```

### Imports

```{r background_setup, include=FALSE}
options(width = 120)
# A trick to display the side-effects:
browseDataList <- function(...) cat(eurodata::browseDataList(...)$html, sep='\n')
find <- function(...) cat(eurodata::find(...)$report, sep='\n')
eurostat::clean_eurostat_cache() # in case repeated evaluation
```

```{r import_demo, warning=FALSE}
x <- importData('nama_10_a10')  # actual dataset
str(x)
head(x,10)
# Friendly error if a wrong data code (similar for importLabels):
tryCatch(importData('nama_10_a10_XXX'),
         error = function(e) cat(geterrmessage()))
y <- importDataList()  # matadata
colnames(y)
str(y[y$Code=='nama_10_a10',])  # matadata on x
z <- importLabels('geo')
head(z,10)
```

### Search

```{r find_demo}
find(bop, its)
find(bop,-ybk,its)
```

```{r browse_demo, results='asis'}
browseDataList(grepl('bop',Code) & grepl('its',Code))
```

## Speed demo

eurodata::**importData** compared to `r lnk('eurostat','get_eurostat')`:

```{r speed_demo}
y <- importDataList()  # matadata

d <- unique(y[y$Type=='dataset','Code']) # dataset codes

set.seed(1234)  # for replicability

d30 <- sample(d, 30) # a random sample of 30 datasets

get_eurostat <- eurostat::get_eurostat  # package 'eurostat' needs to be installed

FUNS <- c('importData', 'get_eurostat')

timeit <- function(funname, code)
    tryCatch(system.time(get(funname)(code))['elapsed'],
             error = function(e) as.numeric(NA)) # due to errors in some 'eurostat' package imports

compileInfo <-
    function(code) {
        FUNS2 <- if (sample(c(TRUE,FALSE), 1))
            FUNS else rev(FUNS) # so that the execution order is random
        res <- data.frame(code,
                          timeit(FUNS2[1], code),
                          timeit(FUNS2[2], code))
        colnames(res) <- c('Data code name',FUNS2)
        res
    }
```

```{r, message=FALSE, warning=FALSE, results='hide'}
L <- lapply(d30, compileInfo)
```

```{r}
Res <- do.call(rbind, L)

Res2 <- within(Res,
       ratio <- get_eurostat/importData)

row.names(Res2) <- NULL # to eliminate the visual noise

Res2 # lower = faster (in seconds)
# How many times eurodata::importData is faster on average?

mean(Res2$ratio, na.rm=TRUE)
median(Res2$ratio, na.rm=TRUE)
```
